---
title: "BU Biostats Faculty Collaboration Network"
author: "Dan Posner"
date: "October 27, 2016"
output: html_document
---

## Package loading
```{r setup, results='hide', message=F}
ipak <- function(pkg){
  new.pkg <- pkg[!(pkg %in% installed.packages()[, "Package"])]
  if (length(new.pkg)) 
    install.packages(new.pkg, dependencies = TRUE)
  sapply(pkg, require, character.only = TRUE)
}

packages <- c("data.table","statnet", "ndtv",
              "scatterplot3d","htmlwidgets","tsna")
ipak(packages)
```

## Data input
We'll read in the dataset I made of pubmed articles with at least 2 biostat faculty co-authors.  
The data matrix is Article X Faculty, with two header columns (article name and year).  
```{r }
# Read in dataset
DT <- fread('data/edge_dat.csv', header=T)
head(DT)
# Remove article title
DT <- DT[,title:=NULL]
```

## Create Adjacency Matrices for each year
```{r}
# Yearly dataset of shared articles
DTlist <- list()
# List of weighted adjacency matrices for each year
Adjlist <- list()
# List of unweighted adjacency matrices for each year
unweighted <- list()

for (i in 1:length(unique(DT$year))) {
  # Subset data.table by year
  DTlist[[i]] = DT[year==unique(DT$year)[i]]
  # remove year column
  DTlist[[i]][,year:=NULL]

  # Create Adjacency matrix 
  #  i.e. we want entries in this matrix to be
  #  dot products of two faculty's authorship
  Adjlist[[i]] = as.matrix(t(DTlist[[i]])) %*% 
                 as.matrix(DTlist[[i]])
  
  # Create unweighted adjacency matrix
  unweighted[[i]] = Adjlist[[i]]
  # Set edge weights > 1 to 1
  unweighted[[i]][unweighted[[i]]>1] = 1
}
```

## Creating a dynamic (i.e. temporal) co-authorship network  
First we create one network per year.
```{r}
#   One network per time point (year)
netlist <- lapply(unweighted, function(x) 
            as.network(x, directed = F))
```
Then we create a dynamic network from the list of static networks
```{r}
# Create dynamic network from list of static nets
minyear = min(DT$year) # starts in 2002
maxyear = max(DT$year) # ends in 2016
tnet <- networkDynamic(network.list=netlist,
            onsets=(minyear-1):(maxyear-1),
            termini=minyear:maxyear,
            create.TEAs=T) # add vertex names
```

## Plots and animations of network changes over time
```{r, message=F}
compute.animation(tnet)

# Time prism projection plot (set displayed slices with at=times)
timePrism(tnet,at=c(2002,2009,2015),
          displaylabels=TRUE,planes = TRUE,
          label.cex=0.5, angle=35,
          displayisolates=F)
```

# Filmstrip plot
```{r}
filmstrip(tnet,displaylabels=FALSE)
```

# Network changes over time
```{r}
render.d3movie(tnet,output.mode = 'htmlWidget', displaylabels=T)
```

# Timeline
If we want to focus solely on the relative durations of collaborations, we can view the dynamics as a timeline by plotting the active spells of edges and vertices. This plot hides network structure and connectivity.
```{r}
timeline(tnet, displaylabels=T)
```

# Proximity timeline
In this view vertices are positioned vertically by their geodesic distance proximity. This means that changes in network structure deflect the vertices' lines into new positions, attempting to keep closely-tied vertices as neighbors. The edges are not shown at all in this view.
```{r}
proximity.timeline(tnet,default.dist=6,
          mode='sammon',labels.at=2009,vertex.cex=4)
```

# Data format of a dynamic network
Have any faculty published together for at least 5 consecutive years?
```{r}
dat=as.data.frame(tnet)
head(dat)
dat[dat$duration>=5,]
```

## Calculate some basic network statistics
How many edges form at each time step for the collaboration network? Conveniently, the values are returned as a time series, plot already knows how to handle them.
```{r}
tEdgeFormation(tnet)
plot(tEdgeFormation(tnet))
```

## Transitivity over time (artificial since didn't use hyperedges)
```{r}
plot(tSnaStats(tnet, 'gtrans'))
```

## Mean degree over time
```{r}
plot(tErgmStats(tnet,'meandeg'),
     main='Mean degree of co-authorship, 1 year interval')
```

## Co-authorship duration from 2002-2016
```{r}
summary( edgeDuration(tnet) )
hist(edgeDuration(tnet, subject='spells'))
```

## Co-authorship duration in two 7-year periods: 2003-2009 vs 2010-2016
```{r}
summary( edgeDuration(tnet,start=2003,end=2009))
summary( edgeDuration(tnet,start=2010,end=2016))
hist(edgeDuration(tnet,start=2003,end=2009, subject='spells'))
hist(edgeDuration(tnet,start=2010,end=2016, subject='spells'))
```

## Most active interdepartmental co-authors
```{r}
vertexDuration(tnet)
hist(vertexDuration(tnet))
```

## Collapsing the dynamic (temporal) network
```{r}
aggnet = network.collapse(tnet, rm.time.info=F, rule='latest')
slice.par<-list(start=2002,end=2016,interval=15, 
                aggregate.dur=15,rule='latest')
compute.animation(tnet,animation.mode='MDSJ',
                  slice.par=slice.par)
# extract the coords so we can do a static plot
coords<-cbind(get.vertex.attribute.active(tnet, 
                                          'animation.x',at=2002),
              get.vertex.attribute.active(tnet, 
                                          'animation.y',at=2002))
plot(aggnet,coord=coords,
     edge.lwd='activity.duration', edge.col='#00000055',
     edge.label='activity.duration',edge.label.col='blue',
     edge.label.cex=0.5,main='Edge weights ignored', displaylabels=T)
```
